---
title: "NSAA analysis"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F)
library(caret)
library(dplyr)
library(glmnet)
library(car)
library(lme4)
library(lmerTest)
library(MuMIn)
library(ggplot2)
library(ggbeeswarm)
```

Define constants:

```{r constants}

# random seed:
student.number = 4645251

# train/test split:
test.split = 0.2

# cross-validation:
val.split = 0.1
nrepeat = 10
```


! IMPORTANT !

Please run "data_cleaning.R" before running this notebook to extract the data 
until LoA, which is needed to go on with the analysis.

Load the data until LoA:

```{r data-cleaning}

data = read.csv('../data/data_up_to_LoA.csv')
# View(data)

# we use complete cases for analysis, rows with missing NSAA_tot are discarded:
cat('full data rows:', nrow(data), '\n')
data.cc = data[complete.cases(data$NSAA_tot),]
data.cc = data.cc[, !(names(data.cc) %in% c('LoA', 'age_LoA_pat', 'age_LoA_hos', 'tenmtr_veloc', 'X6MWT'))]
data.cc$TREAT = as.factor(data.cc$TREAT)
rownames(data.cc) = NULL
cat('complete cases rows:', nrow(data.cc), '\n')
```

How many patients do we now have with complete cases?
```{r npatients}

npatients = length(unique(data.cc$Patient_ID))
cat('number of patients (=groups):', npatients)
```

We first make an initial 80%/20% training/test split. We will evaluate the performance of
our final mixed model with the selected predictors from the elastic net stage on
that test set.

```{r data-split}

set.seed(student.number)

test.patients = sample(
  unique(data.cc$Patient_ID)
  , size=round(npatients * test.split)
  , replace = FALSE
)

df.test = data.cc[(data.cc$Patient_ID %in% test.patients),]
df.train = data.cc[!(data.cc$Patient_ID %in% test.patients),]

trainpatients = unique(df.train$Patient_ID)

ntrain = nrow(df.train)
ntest = nrow(df.test)
cat('training set visits:', ntrain,'\n')
cat('training set patients:', length(trainpatients),'\n','\n')
cat('test set visits:', ntest,'\n')
cat('test set patients:', length(test.patients))
```

We are going to use grouped cross-validation since there is clustering (repeated measures).


First, we want to get an idea of the average # of obs per patient:

```{r patients-split}

ntest.pat = length(test.patients)
ntrain.pat = npatients - ntest.pat
cat(round(nrow(df.train) / ntrain.pat, 3), 'avg obs / patient')
```

Since our dataset is relatively small, we will aim for a 90/10 training/validation
split.

In short: we want to hold out 10% for validation, and 10% of 73 patients is 7.13
patients on average to hold out.
( This means we will hold out 10% of 209 observations for validation = 20.9 obs.
20.900 / 2.931 = 7.13 groups
Therefore, we want to hold out 7.13 patients on average in each train/val split.)

Which means we will use k=10 folds in ``groupKFold()``.

Adjust the validation split constant as you wish below to get # folds:

```{r cv-folds}

cat(paste0('validation split of ', val.split*100, '% gives ', round(1/val.split,1), ' folds'))
```

Note that the number of visits of our patients has some variance, some patients 
have up to 8 visits. Thus, some validation splits might have much more than
20.9 observations.

For this reason, we repeat the cross-validation 10 times to correct for the variance
induced in the CV RMSE by the variance in the number of visits per patient.


Define predictor matrix and outcome vector:

```{r xy-data-split}

X = subset(df.train, select=-c(Patient_ID, NSAA_tot))
X$TREAT = as.numeric(X$TREAT) - 1  # need to make TREAT binary to calculate inner products
y = df.train$NSAA_tot
```

Build hyperparameter grid:

```{r make-tune-grid}

nlambda = 100  # default in glmnet

compute_lambda_max = function(X.fold, y.fold, alpha) {
  # inspired by "Regularization paths for generalized linear models via coordinate descent" (Friedman, Hastie and Tibshirani, 2010)
  n = length(y.fold)
  # below, we standardise X, the same way glmnet does internally
  # during the CV, for consistency (we cross-checked with glmnet function).
  # glmnet uses 1/n standard deviation, so we apply that too:
  X.stand = scale(X.fold, center = TRUE, scale = TRUE) * sqrt(n / (n-1))
  abs.inner.prods = abs(crossprod(X.stand, y.fold))
  lambda.max = max(abs.inner.prods) / (n * alpha)
}

# compute_lambda_max_glmnet = function(X.fold, y.fold, alpha) {
#   fit = glmnet(
#     X.fold
#     , y.fold
#     , alpha=alpha
#     , family='gaussian'
#     , standardize=TRUE
#     , nlambda=nlambda
#   )
#   max(fit$lambda)
# }

# create folds for each repeat of the nrepeat CVs:
nfolds = as.integer(round(1/val.split))
random.seeds = seq(from=student.number, to=student.number + nrepeat - 1)
all.folds = lapply(random.seeds, function(seed) {
    set.seed(seed)
    groupKFold(df.train$Patient_ID, k=nfolds)
  })

# compute lambda.max for every training split (nfolds*nrepeat splits)
alphas = seq(from=0, to=1, by=0.1)

all.lmax = lapply(alphas[-1], function(alpha) {
  unlist(lapply(all.folds, function(cv.folds) {
    lapply(cv.folds, function(fold.indices) {
        X.fold = X[fold.indices,]
        y.fold = y[fold.indices]
        lmax.manual = compute_lambda_max(X.fold, y.fold, alpha)
        # lmax.glmnet = compute_lambda_max_glmnet(X.fold, y.fold, alpha)
        # print(cat('manual:',lmax.manual,'glmnet:',lmax.glmnet))
      })
    }))
  })
names(all.lmax) = alphas[-1]

# compute the lambda sequences for each training split (nfolds*nrepeat splits):
compute_lambdas = function(alpha) {
  
  # little math trick here, we need log10 because we'll create a sequence
  # of lambdas on the log scale.
  # we follow the advice from glmnet docs where lambda.min/lambda.max should be
  # 0.01 in the high-dimensional case, thus:
  if (alpha!=0) {
    lmax.vals = all.lmax[[as.character(alpha)]]
    lmax.max = max(lmax.vals)
    lmax.min = min(lmax.vals)
    
    grid = expand.grid(
      alpha=alpha
      , lambda=10^seq(
          from=log10(lmax.max), to=(log10(lmax.min)-2), length.out=nlambda
        )
      )
  } else {
    # Ridge case needs to be handled separately, as lambda.max is undefined when alpha is 0
    # range of lambdas is set arbitrarily
    grid = expand.grid(
      alpha=0
      , lambda=10^seq(
          from=3, to=1, length.out=nlambda
        )
      )
  }
}


# final tuning grid
tune.grid = do.call(rbind, lapply(alphas, compute_lambdas))
```

Run CV:

```{r grouped-cv}

# we don't need Patient_ID anymore from here on
df.cv = subset(df.train, select=-Patient_ID)

elastic_net_grouped_CV = function(i) {
  
  folds = all.folds[[i]]
  
  train.control = trainControl(
    method='cv'
    , index=folds
    , verboseIter = F
  )

  elastic_net = train(
    NSAA_tot ~ .
    , data=df.cv
    , method='glmnet'
    , standardize = TRUE
    , trControl=train.control
    , tuneGrid=tune.grid
  )
}

models = lapply(seq_along(random.seeds), elastic_net_grouped_CV)

# average rmse's across nrepeat CV repetitions:
avg.rmse.grid = do.call(rbind, lapply(models, function(model) {
  model$results[, c('alpha', 'lambda', 'RMSE')]
})) %>%
  group_by(alpha, lambda) %>%
  summarise(
    rmse.avg = mean(RMSE, na.rm=TRUE)
    , rmse.se = sd(RMSE, na.rm=TRUE) / sqrt(nrepeat)  # SE of MC estimator of rmse 
    , .groups='drop'
  )

one_SE = function(grid)
  # inspired by caret::oneSE()
  {
    grid = grid %>%
      arrange(desc(lambda))
    best.index = which.min(grid$rmse.avg)
    thresh = grid$rmse.avg[best.index] + grid$rmse.se[best.index]
    candidates = which(grid$rmse.avg <= thresh)
    best.iter = min(candidates)
    list(lambda=grid$lambda[best.iter], cv.rmse=grid$rmse.avg[best.iter]
         , rmse.se=grid$rmse.se[best.iter])
}

best.pairs = avg.rmse.grid %>%
  group_by(alpha) %>%
  group_map(~ {
    best.pair = one_SE(.x)
    data.frame(
      alpha=.y$alpha
      , best.lambda=best.pair$lambda
      , rmse=best.pair$cv.rmse
      , rmse.se=best.pair$rmse.se
    )}) %>%
  do.call(rbind, .)

```

Note: the "There were missing values in resampled performance measures." warnings
are normal, when the lambda being fitted with is equal or greater than lambda.max
in some folds, the r-squared is NaN since the variance of the predictions is zero (intercept model).


Now that we have our best (alpha, lambda) pairs, we refit an elastic net to
our entire training set with each of these pairs to obtain Table 3 of our report:

```{r best-tuned-models}

all.nonzero.preds = list()

for (i in 1:nrow(best.pairs)) {
  
  alpha = best.pairs$alpha[i]
  lambda = best.pairs$best.lambda[i]
  rmse = best.pairs$rmse[i]
  rmse.se = best.pairs$rmse.se[i]
  
  X.scaled = scale(X, center = TRUE, scale = TRUE)
  
  fit = glmnet(
    X.scaled
    , y
    , family='gaussian'
    , alpha=alpha
    , lambda=lambda
  )
  coefs = coef(fit)[-1,]  # drop the intercept, we don't need it here
  nonzero.preds = names(which(coefs != 0))
  all.nonzero.preds[[i]] = nonzero.preds
  
  # check condition number
  mat.predictors = as.matrix(X[, nonzero.preds])
  cond.num = kappa(t(mat.predictors) %*% mat.predictors)
  
  print(paste('pair number', i, ' (alpha='
        , alpha
        , ', lambda='
        , round(lambda, 3)
        , '): '
        , '# non-zero predictors: '
        , length(nonzero.preds)
        , ', kappa = '
        , round(cond.num, 3)
        , ', CV RMSE = '
        , round(rmse, 3)
        , ', RMSE SE = '
        , round(rmse.se, 3)
  ))
}
```

The retained (non-zero) predictors of pair number 'n' are stored in
``all.nonzero.preds[[n]]``.

```{r save-stage1-proteins}

# pair with the lowest condition number and smallest predictor set:
# all.nonzero.preds[[8]]
best.index = 8
proteins = list(proteins=all.nonzero.preds[[best.index]])
write.csv(proteins, 'stage1_proteins.csv', row.names = F)
df = read.csv('../data/stage1_proteins.csv')
(stage1.proteins = df$proteins)
```

Fit elastic net with best parameters:

```{r stage2-data-preprocess}

# Model with covariates
continuous.cols = setdiff(names(df.train), c("NSAA_tot", "Patient_ID", "TREAT"))
train.mean = sapply(df.train[, continuous.cols], mean)
train.sd = sapply(df.train[, continuous.cols], sd)
train.sd[train.sd == 0] = 1

# scale
df.train.scaled = df.train
df.train.scaled[, continuous.cols] = scale(df.train[, continuous.cols], center = train.mean, scale = train.sd)

df.test.scaled = df.test
df.test.scaled[, continuous.cols] = scale(df.test[, continuous.cols], center = train.mean, scale = train.sd)

predictors.with.cov = c("age", "TREAT", setdiff(stage1.proteins, c("age", "TREAT")))
```

Stage 2 C-EN (cluster-informed elastic net):

```{r stage2-c-en}

pat.counts = df.train %>% count(Patient_ID)

# build design matrix of random intercepts:
Z = matrix(0, ntrain, ntrain.pat)
i = 1
for (j in 1:ntrain.pat) {
  n.visits = pat.counts$n[j]
  Z[i:(i+n.visits-1), j] = 1
  i = i + n.visits
}

# proxy matrix of random effects covariance
M = log(ntrain.pat) * diag(ntrain.pat)
Pz = solve( diag(ntrain) + Z %*% M %*% t(Z) )

# cholesky decomp & define new y and X
# chol() returns upper triangular factor
U = chol(Pz)

y = df.train.scaled$NSAA_tot
X = subset(df.train.scaled, select=predictors.with.cov)
X$TREAT = as.numeric(X$TREAT) - 1
X = as.matrix(X)
y.tilde = U %*% y
X.tilde = U %*% X

# elastic net regularisation
alpha = best.pairs$alpha[best.index]
lambda = best.pairs$best.lambda[best.index]

fit = glmnet(X.tilde, y.tilde, alpha=alpha, lambda=lambda)

coef.2 = coef(fit)[-1,]
nonzero.preds.2 = names(which(coef.2 != 0))
write.csv(nonzero.preds.2, '../data/final_proteins.csv', row.names = F)
```

Finally, fit stage 3 LMM and eval on test set:

```{r stage3-lmm-eval}

# fit the random intercept LMM
predictors.2 = c(
  "age", "TREAT", setdiff(nonzero.preds.2, c("age", "TREAT"))
)
formula.2 = as.formula(
  paste("NSAA_tot ~", paste(predictors.2, collapse = " + ")
        , "+ (1 | Patient_ID)")
)

lmm = lmer(formula.2, data = df.train.scaled)
summary(lmm)
# conditional & marginal R²:
r.squaredGLMM(lmm)

# test set prediction
y.pred = predict(
  lmm,
  newdata = df.test.scaled,
  allow.new.levels = TRUE
)

y.test = df.test.scaled$NSAA_tot
rmse.test = sqrt(mean((y.test - y.pred)^2))
rmse.test

r2.test = 1 - sum((y.test - y.pred)^2) / sum((y.test - mean(y.test))^2)
r2.test
```

Test for RANDOM INTERCEPT + SLOPE:

```{r rslope-lmm}

rslope.formula = as.formula(
  paste("NSAA_tot ~", paste(predictors.2, collapse = " + ")
        , "+ (1 + age | Patient_ID)")
)

lmm.rslope = lmer(rslope.formula, data = df.train.scaled)
summary(lmm.rslope)

anova(lmm, lmm.rslope)
```

Check: not worth it to add a random slope on age.

RMSE per-patient:

```{r per-pat-rmse}

per_patient_rmse = function(pat) {

    df.pat = df.test.scaled[df.test.scaled$Patient_ID == pat,]
    y.test = df.pat$NSAA_tot

    y.pred = predict(
      lmm,
      newdata = df.pat,
      allow.new.levels = TRUE
    )
    rmse = round( sqrt( mean( (y.test - y.pred)^2 ) ), 3)
    
    c(
      Patient_ID=pat
      , nobs=nrow(df.pat)
      , RMSE = rmse
    )
}

test.patients = sort(test.patients)
test.results = as.data.frame(t(sapply(test.patients, per_patient_rmse)))
test.results = test.results[order(test.results$nobs),]
rownames(test.results) = NULL
test.results

# png(filename = '../plots/per_patient_RMSE.png', res=150, width=1200, height=900)
ggplot(test.results, aes(x=factor(nobs), y=RMSE, color=nobs)) +
  labs(x='Patient visits', y='Patient RMSE') +
  geom_beeswarm(size=4) +
  geom_hline(yintercept=rmse.test, lty=2, color='grey50', linewidth=0.6) +
  scale_color_gradient(low='cyan3', high='darkblue') +
  ggtitle('Per-patient test RMSE') +
  theme(legend.position='none', plot.title=element_text(face='bold')) +
  annotate(
    'text',
    x=max(test.results$nobs)+1,
    y=rmse.test,
    label='Full-sample test RMSE',
    color='grey40',
    hjust=1.1,
    vjust=-0.6,
    size=3.3
)
# dev.off()
```

Check LMM assumptions:

```{r lmm-diag-plots}

fitted.val = fitted(lmm)
raw.resid = resid(lmm)
jack.resid = rstudent(lmm)

# obs vs fitted
plot(fitted.val, df.train$NSAA_tot
     , xlab = "Fitted value"
     , ylab = "NSAA_tot"
     , main='Observed NSAA_tot vs fitted values')

# qq plots
# JACK:
# png(filename = '../plots/jack_qq.png', res=150, width=1200, height=900)
qqnorm(jack.resid, main = 'Jackknife conditional residuals Q-Q plot')
qqline(jack.resid, col = "red")
# dev.off()
# RAW:
qqnorm(raw.resid, main = 'Raw conditional residual Q-Q plot')
qqline(raw.resid, col = "red")

# histograms
# JACK:
hist(jack.resid, breaks = 30, main = 'Jackknife conditional residuals histogram', 
     xlab='Jackknife conditional residual')
# RAW:
hist(raw.resid, breaks = 30, main = 'Raw conditional residuals histogram', 
     xlab='Raw conditional residual')

# resid vs fitted
# RAW:
plot(fitted.val, raw.resid,
     xlab = "Fitted value",
     ylab = "Raw conditional residual",
     main = "Raw conditional residuals vs fitted values")
abline(h = 0, col = "red")
lines(lowess(fitted.val, raw.resid),
      col = "blue", lwd = 2)

# JACK:
# png(filename = '../plots/jack_vs_fitted.png', res=150, width=1200, height=900)
plot(fitted.val, jack.resid,
     xlab = "Fitted value",
     ylab = "Residual",
     main = "Jackknife conditional residuals vs fitted values")
abline(h = 0, col = "red")
lines(lowess(fitted.val, jack.resid),
      col = "blue", lwd = 2)
# dev.off()



# random effects diagnostic plots:
ran.int = ranef(lmm)$Patient_ID
jack.resid = rstudent(lmm)

dfs = lapply(trainpatients, function(p){
      rows = rownames(df.train[df.train$Patient_ID == p,])
      pat.resid = unname(jack.resid[rows])
      pat.ranint = rep(ran.int[as.character(p),], length(pat.resid))
      data.frame(residual=pat.resid, ranint=pat.ranint)
    }
)
df.resid.ranint = bind_rows(dfs)

# png(filename = '../plots/jack_vs_ranint.png', res=150, width=1200, height=900)
ggplot(df.resid.ranint, aes(x=ranint, y=residual)) +
  labs(x='Random intercept', y='Jackknife residual') +
  geom_point(size=2, shape=1) +
  geom_smooth(method='loess', se=F, fullrange=F
              , method.args=list(degree =1, family = 'gaussian')) +
  scale_color_gradient(low='cyan3', high='darkblue') +
  ggtitle('Jackknife conditional residuals vs random effects') +
  theme(legend.position='none', plot.title=element_text(face='bold'))
# dev.off()

# png(filename = '../plots/jack_vs_ranint.png', res=150, width=1200, height=900)
plot(df.resid.ranint$ranint, df.resid.ranint$residual,
     xlab = "Random intercept",
     ylab = "Residual",
     main = "Jackknife conditional residuals vs random intercepts")
abline(h = 0, col = "red")
lines(lowess(df.resid.ranint$ranint, df.resid.ranint$residual),
      col = "blue", lwd = 2)
# dev.off()

hist(ran.int$`(Intercept)`, breaks=20, xlab='Random intercept'
     , main='Histogram of random intercepts')

# png(filename = '../plots/ranint_hist.png', res=150, width=1200, height=900)
ggplot(ran.int, aes(x=`(Intercept)`)) + 
  labs(x='Random intercept', y='Frequency') +
  geom_histogram(binwidth = 1, colour='black', fill='grey') +
  geom_density(aes(y = after_stat(density * nrow(ran.int) * 1)),
               alpha=.2, fill="#FF6666") +
  ggtitle('Histogram of random intercepts')
# dev.off()
```


Rank coefficients:

```{r rank-coefs}

coefs.model = summary(lmm)$coefficients
coefs.model = coefs.model[rownames(coefs.model) != "(Intercept)", ]
coefs.model

ranked_coefs_est = data.frame(
  variable = rownames(coefs.model),
  estimate = coefs.model[, "Estimate"],
  row.names = NULL
)
ranked_coefs_est = ranked_coefs_est[order(abs(ranked_coefs_est$estimate), decreasing = TRUE), ]
ranked_coefs_est
```


Check for influential patients:

```{r influential-patients}
library(lme4)
library(influence.ME)

model_lme4 <- lme4::lmer(
  NSAA_tot ~ X10440.26 + X11277.23+ X12549.33+ X13114.50+ X15386.7+ X15559.5+  X19630.2+  X3220.40+  X3709.4+  X4929.55+  X5000.52+  X5483.1+   X5542.22+  X5864.10+  X7970.315+ X8080.24+ X8300.82+  X8484.24+ X9851.9 +
    age + TREAT + (1 | Patient_ID), data = df.train.scaled)

infl <- influence(model_lme4, group = "Patient_ID")

cd <- cooks.distance(infl)
plot(cd, type = "h",
     ylab = "Cook's distance",
     xlab = "Patient",
     main = "Patient-level influence")

plot(cd, main = "Cooks Distance for Influential Obs");abline(h = 4/length(unique(df.train.scaled$Patient_ID)), lty = 2, col = "steelblue")
```