---
title: "NSAA analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(caret)
library(dplyr)
library(glmnet)
library(car)
library(lme4)
library(lmerTest)
library(MuMIn)
library(ggplot2)
library(ggbeeswarm)
```

Define constants:

```{r}
# random seed:
student.number = 4645251

# train/test split:
test.split = 0.2

# cross-validation:
val.split = 0.1
nrepeat = 10
```

Load the data:

```{r data-cleaning}
data = read.csv('../data/data_up_to_LoA.csv')
if('X' %in% names(data)) data = subset(data, select=-X)

# using complete cases for analysis:
cat('full data rows:', nrow(data), '\n')
data_cc = data[complete.cases(data$NSAA_tot),]
data_cc = data_cc[, !(names(data_cc) %in% c('LoA', 'age_LoA_pat', 'age_LoA_hos', 'tenmtr_veloc', 'X6MWT'))]
data_cc$TREAT = as.factor(data_cc$TREAT)
rownames(data_cc) = NULL
cat('complete cases rows:', nrow(data_cc), '\n')
```

Quick look at out data:

```{r}
head(data_cc)
```

How many patients do we now have with complete cases?
```{r}
n.patients = length(unique(data_cc$Patient_ID))
cat('number of patients (=groups):', n.patients)
```

We first make an initial training/test split. We will evaluate the performance of
our final mixed model with the selected predictors from the elastic net stage on
that test set.

```{r}
set.seed(student.number)
test.patients = sample(
  unique(data_cc$Patient_ID)
  , size=round(n.patients * test.split)
  , replace = FALSE
)
df.test = data_cc[(data_cc$Patient_ID %in% test.patients),]
df.train = data_cc[!(data_cc$Patient_ID %in% test.patients),]

train.patients = unique(df.train$Patient_ID)

(n.train = nrow(df.train))
(n.test = nrow(df.test))
```


We are going to use grouped cross-validation since we have repeated measurements.

Here's the rationale for the number of groups held out in the grouped CV:
We have 73 patients (groups), and 209 observations.

First, we want to get an idea of the average # of obs per patient:
```{r}
n.test.patients = length(test.patients)
n.train.patients = n.patients - n.test.patients
cat(round(nrow(df.train) / n.train.patients, 3), 'avg obs / patient')
```

Since our dataset is relatively small, we will aim for a 90/10 training/validation
split.

In short: we want to hold out 10% for validation, and 10% of 73 patients is 7.3
patients on average to hold out.
( This means we will hold out 10% of 209 observations for validation = 20.9 obs.
20.900 / 2.863 = 7.300 groups
Therefore, we want to hold out 7.3 patients on average in each train/val split. )

Which means: we will use k=10 folds in ``groupKFold()``.

Adjust the validation split as you wish below to get # folds:
```{r}
cat(paste0('validation split of ', val.split*100, '% gives ', round(1/val.split,1), ' folds'))
```

Note that the number of visits of our patients has some variance, some patients 
have until 8 visits. Thus, some validation splits might have much more than
20.9 observations.

For this reason, we repeat the cross-validation 10 times to correct for the variance
induced in the CV MSE by the variance in the number of visits per patient.

```{r make-tune-grid}

nlambda = 100  # default in glmnet
X = subset(df.train, select=-c(Patient_ID, NSAA_tot))
X$TREAT = as.numeric(X$TREAT) - 1  # need to make TREAT binary to calculate inner products
y = df.train$NSAA_tot

compute_lambda_max = function(X.fold, y.fold, alpha) {
  # inspired by "Regularization paths for generalized linear models via coordinate descent" (Friedman, Hastie and Tibshirani, 2010)
  n = length(y.fold)
  # below, we standardise X, the same way glmnet does internally
  # during the CV, for consistency (we cross-checked with glmnet function).
  # glmnet uses 1/n standard deviation, so we follow that here too
  X.stand = scale(X.fold, center = TRUE, scale = TRUE) * sqrt(n / (n-1))
  abs.inner.prods = abs(crossprod(X.stand, y.fold))
  lambda.max = max(abs.inner.prods) / (n * alpha)
}

# compute_lambda_max_glmnet = function(X.fold, y.fold, alpha) {
#   fit = glmnet(
#     X.fold
#     , y.fold
#     , alpha=alpha
#     , family='gaussian'
#     , standardize=TRUE
#     , nlambda=nlambda
#   )
#   max(fit$lambda)
# }

# create folds for each repeat of the nrepeat CVs:
nfolds = as.integer(round(1/val.split))
random.seeds = seq(from=student.number, to=student.number + nrepeat - 1)
all.folds = lapply(random.seeds, function(seed) {
    set.seed(seed)
    groupKFold(df.train$Patient_ID, k=nfolds)
  })

# compute lambda.max for every training split (nfolds*nrepeat splits)
alphas = seq(from=0, to=1, by=0.1)

# scaled.folds = lapply(all.folds, function(cv.folds) {
#     lapply(cv.folds, function(fold.indices) {
#         X.fold = X[fold.indices,]
#         y.fold = y[fold.indices]
#         n = nrow(y.fold)
#         # glmnet uses 1/n SD for normalisation
#         X.scaled = scale(X.fold, center = TRUE, scale = TRUE) * sqrt(n / (n-1))
#         y.scaled = scale(y.fold, center = TRUE, scale = TRUE) * sqrt(n / (n-1))
#         list(X=X.scaled, y=y.scaled)
#       })
#   })

all.lmax = lapply(alphas[-1], function(alpha) {
  unlist(lapply(all.folds, function(cv.folds) {
    lapply(cv.folds, function(fold.indices) {
        X.fold = X[fold.indices,]
        y.fold = y[fold.indices]
        lmax.manual = compute_lambda_max(X.fold, y.fold, alpha)
        # lmax.glmnet = compute_lambda_max_glmnet(X.fold, y.fold, alpha)
        # print(cat('manual:',lmax.manual,'glmnet:',lmax.glmnet))
      })
    }))
  })
names(all.lmax) = alphas[-1]

# compute the lambda sequences for each training split (nfolds*nrepeat splits):
compute_lambdas = function(alpha) {
  
  # little math trick here, we need log10 because we'll create a sequence
  # of lambdas on the log scale.
  # we follow the advice from glmnet docs where lambda.min/lambda.max should be
  # 0.01 in the high-dimensional case, thus:
  if (alpha!=0) {
    lmax.vals = all.lmax[[as.character(alpha)]]
    lmax.max = max(lmax.vals)
    lmax.min = min(lmax.vals)
    
    grid = expand.grid(
      alpha=alpha
      , lambda=10^seq(
          from=log10(lmax.max), to=(log10(lmax.min)-2), length.out=nlambda
        )
      )
  } else {
    # Ridge case needs to be handled separately, as lambda.max is undefined when alpha is 0
    # range of lambdas is set arbitrarily
    grid = expand.grid(
      alpha=0
      , lambda=10^seq(
          from=3, to=1, length.out=nlambda
        )
      )
  }
}

tune.grid = do.call(rbind, lapply(alphas, compute_lambdas))
```

Run CV:

```{r grouped-cv}

# we don't need Patient_ID anymore from here on
df_train_PID= df.train
df.train = subset(df.train, select=-Patient_ID)

elastic_net_grouped_CV = function(i) {
  
  folds = all.folds[[i]]
  
  train.control = trainControl(
    method='cv'
    , index=folds
    , verboseIter = F
  )

  elastic_net = train(
    NSAA_tot ~ .
    , data=df.train
    , method='glmnet'
    , standardize = TRUE
    , trControl=train.control
    , tuneGrid=tune.grid
  )
}

models = lapply(seq_along(random.seeds), elastic_net_grouped_CV)

# average rmse's across nrepeat CV repetitions:
avg.rmse.grid = do.call(rbind, lapply(models, function(model) {
  model$results[, c('alpha', 'lambda', 'RMSE')]
})) %>%
  group_by(alpha, lambda) %>%
  summarise(
    rmse.avg = mean(RMSE, na.rm=TRUE)
    , rmse.se = sd(RMSE, na.rm=TRUE) / sqrt(nrepeat)  # SE of MC estimator of rmse 
    , .groups='drop'
  )

one_SE = function(grid)
  # inspired by caret::oneSE()
  {
    grid = grid %>%
      arrange(desc(lambda))
    best.index = which.min(grid$rmse.avg)
    thresh = grid$rmse.avg[best.index] + grid$rmse.se[best.index]
    candidates = which(grid$rmse.avg <= thresh)
    best.iter = min(candidates)
    list(lambda=grid$lambda[best.iter], cv.rmse=grid$rmse.avg[best.iter])
}

best.pairs = avg.rmse.grid %>%
  group_by(alpha) %>%
  group_map(~ {
    best.pair = one_SE(.x)
    data.frame(
      alpha=.y$alpha
      , best.lambda=best.pair$lambda
      , cv.rmse=best.pair$cv.rmse
    )}) %>%
  do.call(rbind, .)

```
Note: the "There were missing values in resampled performance measures." warnings
are normal, when the lambda being fitted with is equal or greater than lambda.max
in some folds, the r-squared is NaN since the variance of the predictions is zero (intercept model).

Diagnose the lambda situation:
```{r}
model.results = models[7][[1]]$results
model.results[model.results$alpha==0.9,]
```


Now that we have our best (alpha, lambda) pairs, we refit an elastic net to
our entire training set with each of these pairs:

```{r}
all.nonzero.params = list()

for (i in 1:nrow(best.pairs)) {
  
  alpha = best.pairs$alpha[i]
  lambda = best.pairs$best.lambda[i]
  cv.rmse = best.pairs$cv.rmse[i]
  
  X.scaled = scale(X, center = TRUE, scale = TRUE)
  
  fit = glmnet(
    X.scaled
    , y
    , family='gaussian'
    , alpha=alpha
    , lambda=lambda
  )
  coefs = coef(fit)[-1,]  # drop the intercept, we don't need it here
  nonzero.params = names(which(coefs != 0))
  all.nonzero.params[[i]] = nonzero.params
  
  # check condition number
  mat.predictors = as.matrix(X[, nonzero.params])
  cond.num = kappa(t(mat.predictors) %*% mat.predictors)
  
  print(paste('pair number', i, ' (alpha='
        , alpha
        , ', lambda='
        , round(lambda, 3)
        , '): '
        , '# non-zero predictors: '
        , length(nonzero.params)
        , ', kappa = '
        , round(cond.num, 3)
        , ', CV RMSE = '
        , round(cv.rmse, 3)
  ))
}
```

The retained (non-zero) predictors of pair number 'n' are stored in
``all.nonzero.params[[n]]``.

```{r}
# pair with the lowest condition number and smallest predictor set:
# all.nonzero.params[[8]]
# proteins = list(proteins=all.nonzero.params[[9]])
# write.csv(proteins, 'final_proteins.csv', row.names = F)
df = read.csv('../data/final_proteins.csv')
selected.proteins = df$proteins
selected.proteins
```

Fit elastic net with best parameters

```{r}
# Model with covariates
num_cols = setdiff(names(df_train_PID), c("NSAA_tot", "Patient_ID", "TREAT"))
train_mean = sapply(df_train_PID[, num_cols], mean)
train_sd = sapply(df_train_PID[, num_cols], sd)
train_sd[train_sd == 0] = 1

# scale
df.train_scaled = df_train_PID
df.train_scaled[, num_cols] = scale(df_train_PID[, num_cols], center = train_mean, scale = train_sd)

df.test_scaled = df.test
df.test_scaled[, num_cols] = scale(df.test[, num_cols], center = train_mean, scale = train_sd)

parameters.with.cov = c("age", "TREAT", setdiff(selected.proteins, c("age", "TREAT")))
```


Extension (round 2):
```{r}
pat.counts = df_train_PID %>% count(Patient_ID)

# build design matrix of random effects:
Z = matrix(0, n.train, n.train.patients)
i = 1
for (j in 1:n.train.patients) {
  n.visits = pat.counts$n[j]
  Z[i:(i+n.visits-1), j] = 1
  i = i + n.visits
}

# build proxy ranef matrix:
M = log(n.train.patients) * diag(n.train.patients)
Pz = solve( diag(n.train) + Z %*% M %*% t(Z) )

# cholesky decomp & define new y and X
U = chol(Pz)

y = df.train_scaled$NSAA_tot
X = subset(df.train_scaled, select=parameters.with.cov)
X$TREAT = as.numeric(X$TREAT) - 1
X = as.matrix(X)
y.tilde = U %*% y
X.tilde = U %*% X

# elastic net regularisation
alpha = best.pairs$alpha[8]
lambda = best.pairs$best.lambda[8]
fit = glmnet(X.tilde, y.tilde, alpha=alpha, lambda=lambda)
coef.round2 = coef(fit)[-1,]
nonzero.proteins.round2 = names(which(coef.round2 != 0))

# fit the random intercept LMM
predictors.round2 = c(
  "age", "TREAT", setdiff(nonzero.proteins.round2, c("age", "TREAT"))
)
formula.round2 = as.formula(
  paste("NSAA_tot ~", paste(predictors.round2, collapse = " + ")
        , "+ (1 | Patient_ID)")
)
lmm.round2 = lmer(formula.round2, data = df.train_scaled)
summary(lmm.round2)

# hold out prediction
y.pred.round2 = predict(
  lmm.round2,
  newdata = df.test_scaled,
  allow.new.levels = TRUE
)

y.test = df.test_scaled$NSAA_tot
rmse.test.round2 = sqrt(mean((y.test - y.pred.round2)^2))
rmse.test.round2

r2.test.round2 = 1 - sum((y.test - y.pred.round2)^2) / sum((y.test - mean(y.test))^2)
r2.test.round2
```
save final proteins:
```{r}
# write.csv(nonzero.proteins.round2, '../data/proteins_round2.csv', row.names = F)
```


```{r}
r.squaredGLMM(lmm.round2)
```

WITH RANDOM INTERCEPT + SLOPE:
```{r}
rslope.formula = as.formula(
  paste("NSAA_tot ~", paste(predictors.round2, collapse = " + ")
        , "+ (1 + age | Patient_ID)")
)

lmm.round2.rslope = lmer(rslope.formula, data = df.train_scaled)
summary(lmm.round2.rslope)
```

Model comparison:
```{r}
anova(lmm.round2, lmm.round2.rslope)
```
Check: not worth it to add a random slope on age.

R² per-patient:

```{r}
per_patient_rmse = function(pat) {

    df.pat = df.test_scaled[df.test_scaled$Patient_ID == pat,]
    y.test = df.pat$NSAA_tot

    y.pred = predict(
      lmm.round2,
      newdata = df.pat,
      allow.new.levels = TRUE
    )
    rmse = round( sqrt( mean( (y.test - y.pred)^2 ) ), 3)
    
    c(
      Patient_ID=pat
      , n.obs=nrow(df.pat)
      , RMSE = rmse
    )
}

test.patients = sort(test.patients)
test.results = as.data.frame(t(sapply(test.patients, per_patient_rmse)))
test.results = test.results[order(test.results$n.obs),]
rownames(test.results) = NULL
test.results

# png(filename = '../plots/per_patient_RMSE.png', res=150, width=1200, height=800)
ggplot(test.results, aes(x=factor(n.obs), y=RMSE, color=n.obs)) +
  labs(x='Number of visits of the patient', y='Per-patient RMSE') +
  geom_beeswarm(size=4) +
  geom_hline(yintercept=rmse.test.round2, lty=2, color='grey50', linewidth=0.6) +
  scale_color_gradient(low='cyan3', high='darkblue') +
  ggtitle('Per-patient test RMSE') +
  theme(legend.position='none', plot.title=element_text(face='bold')) +
  annotate(
    'text',
    x=max(test.results$n.obs)+1,
    y=rmse.test.round2,
    label='Full-sample test RMSE',
    color='grey40',
    hjust=1.1,
    vjust=-0.6,
    size=3.3
)
# dev.off()
```


Checking linearity:

```{r}
fitted.val = fitted(lmm.round2)
raw.resid = resid(lmm.round2)
jack.resid = rstudent(lmm.round2)

# obs vs fitted
plot(fitted.val, df.train$NSAA_tot
     , xlab = "Fitted value"
     , ylab = "NSAA_tot"
     , main='Observed NSAA_tot vs fitted values')

# qq plots
# png(filename = '../plots/jack_qq.png', res=150, width=1200, height=900)
qqnorm(jack.resid, main = 'Jackknife conditional residuals Q-Q plot')
qqline(jack.resid, col = "red")
# dev.off()

qqnorm(raw.resid, main = 'Raw conditional residual Q-Q plot')
qqline(raw.resid, col = "red")

# histograms
hist(jack.resid, breaks = 30, main = 'Jackknife conditional residuals histogram', 
     xlab='Jackknife conditional residual')
hist(raw.resid, breaks = 30, main = 'Raw conditional residuals histogram', 
     xlab='Raw conditional residual')

# resid vs fitted
plot(fitted.val, raw.resid,
     xlab = "Fitted value",
     ylab = "Raw conditional residual",
     main = "Raw conditional residuals vs fitted values")
abline(h = 0, col = "red")
lines(lowess(fitted.val, raw.resid),
      col = "blue", lwd = 2)

# png(filename = '../plots/jack_vs_fitted.png', res=150, width=1200, height=900)
plot(fitted.val, jack.resid,
     xlab = "Fitted value",
     ylab = "Jackknife conditional residual",
     main = "Jackknife conditional residuals vs fitted values")
abline(h = 0, col = "red")
lines(lowess(fitted.val, jack.resid),
      col = "blue", lwd = 2)
# dev.off()
```
Here we look at the random effects diagnostic plots:

```{r}
ran.int = ranef(lmm.round2)$Patient_ID
jack.resid = rstudent(lmm.round2)

dfs = lapply(train.patients, function(p){
      rows = rownames(df_train_PID[df_train_PID$Patient_ID == p,])
      pat.resid = unname(jack.resid[rows])
      pat.ranint = rep(ran.int[as.character(p),], length(pat.resid))
      data.frame(residual=pat.resid, ranint=pat.ranint)
    }
)
df.resid.ranint = bind_rows(dfs)

# png(filename = '../plots/jack_vs_ranint.png', res=150, width=1200, height=900)
ggplot(df.resid.ranint, aes(x=ranint, y=residual)) +
  labs(x='Random intercept', y='Jackknife residual') +
  geom_point(size=2, shape=1) +
  geom_smooth(method='loess', se=F, fullrange=F
              , method.args=list(degree =1, family = 'gaussian')) +
  scale_color_gradient(low='cyan3', high='darkblue') +
  ggtitle('Jackknife conditional residuals vs random effects') +
  theme(legend.position='none', plot.title=element_text(face='bold'))
# dev.off()

# png(filename = '../plots/jack_vs_ranint.png', res=150, width=1200, height=900)
plot(df.resid.ranint$ranint, df.resid.ranint$residual,
     xlab = "Random intercept",
     ylab = "Jackknife conditional residual",
     main = "Jackknife conditional residuals vs random intercepts")
abline(h = 0, col = "red")
lines(lowess(df.resid.ranint$ranint, df.resid.ranint$residual),
      col = "blue", lwd = 2)
# dev.off()

hist(ran.int$`(Intercept)`, breaks=20, xlab='Random intercept'
     , main='Histogram of random intercepts')

# png(filename = '../plots/ranint_hist.png', res=150, width=1200, height=900)
ggplot(ran.int, aes(x=`(Intercept)`)) + 
  labs(x='Random intercept', y='Frequency') +
  geom_histogram(binwidth = 1, colour='black', fill='grey') +
  geom_density(aes(y = after_stat(density * nrow(ran.int) * 1)),
               alpha=.2, fill="#FF6666") +
  ggtitle('Histogram of random intercepts')
# dev.off()
```


Rank coefficients:

```{r}
coefs.model = summary(lmm.round2)$coefficients
coefs.model = coefs.model[rownames(coefs.model) != "(Intercept)", ]
coefs.model

ranked_coefs_est = data.frame(
  variable = rownames(coefs.model),
  estimate = coefs.model[, "Estimate"],
  row.names = NULL
)
ranked_coefs_est = ranked_coefs_est[order(abs(ranked_coefs_est$estimate), decreasing = TRUE), ]
ranked_coefs_est
```

```{r}
#plot(NSAA_tot ~ ., data = data_cc);lines(weighted.mean(data_cc$NSAA_tot))
```

```{r}
library(lme4)
library(influence.ME)

model_lme4 <- lme4::lmer(
  NSAA_tot ~ X10440.26 + X11277.23+ X12549.33+ X13114.50+ X15386.7+ X15559.5+  X19630.2+  X3220.40+  X3709.4+  X4929.55+  X5000.52+  X5483.1+   X5542.22+  X5864.10+  X7970.315+ X8080.24+ X8300.82+  X8484.24+ X9851.9 +
    age + TREAT + (1 | Patient_ID), data = df.train_scaled)

infl <- influence(model_lme4, group = "Patient_ID")

cd <- cooks.distance(infl)
plot(cd, type = "h",
     ylab = "Cook's distance",
     xlab = "Patient",
     main = "Patient-level influence")

plot(cd, main = "Cooks Distance for Influential Obs");abline(h = 4/length(unique(df.train_scaled$Patient_ID)), lty = 2, col = "steelblue")


```